Accessing and Analyzing Linguistic Field Data

### 2006-10-31 slides/data.tex
updated with simplified examples from chapter
svn/trunk@3664

\begin{frame}
\frametitle{Tools and Technologies}
\begin{itemize}
\item General-purpose tools
  \begin{itemize}
  \item word processors
  \item spreadsheets
  \item databases
  \end{itemize}
\item Special prupose tools
  \begin{itemize}
  \item Toolbox
  \end{itemize}
\item broader question: data management in support of best practices
\item Bird, Steven and Gary Simons (2003).  Seven Dimensions of Portability
    for Language Documentation and Description, \textit{Language} 79: 557-582.
\end{itemize}
\end{frame}

@@

\begin{verbatim}
<p class=MsoNormal>sleep <span style='mso-spacerun:yes'> </span>[<span
class=SpellE>sli:p</span>]<span style='mso-spacerun:yes'>  </span><b><span
style='font-size:11.0pt'>vi</span></b><span style='mso-spacerun:yes'>  </span><i>a
condition of body and mind ...o:p></o:p></i></p>
\end{verbatim}

@@

\begin{itemize}
\item scan the file, convert into tree object
\item preserves order of fields, gives array and XPath-style access
\item example: calculate average number of fields per entry
\end{itemize}

\begin{verbatim}
>>> from nltk_lite.corpora import toolbox
>>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
>>> sum(len(entry) for entry in lexicon) / len(lexicon)
10
\end{verbatim}
\end{frame}

@@

\scriptsize
\begin{verbatim}
>>> for field in lexicon[3]:
...     print "\\%s %s" % (field.tag, field.text)
\lx kaa
\ps N.M
\cl isi
\ge cooking banana
\gp banana bilong kukim
\sf FLORA
\dt 12/Feb/2005
\ex Taeavi iria kaa isi kovopaueva kaparapasia.
\xp Taeavi i bin planim gaden banana bilong kukim tasol long paia.
\xe Taeavi planted banana in order to cook it.
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> from nltk_lite.etree.ElementTree import Element
>>> for entry in lexicon:
...     for i in range(len(entry)):
...         if entry[i].tag == 'lx':
...             cv_field = Element("cv")
...             cv_field.text = cv(entry[i].text)
...             entry.insert(i+1, cv_field)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Adding New Fields (cont)}
\small
\begin{verbatim}
>>> for field in lexicon[50]:
...     print "\\%s %s" % (field.tag, field.text)
\lx kaeviro
\cv CVVCVCV
\ps V.A
\ge lift off
\ge take off
\gp go antap
\nt used to describe action of plane
\dt 12/Feb/2005
\ex Pita kaeviroroe kepa kekesia oa vuripierevo kiuvu.
\xp Pita i go antap na lukim haus win i bagarapim.
\xe Peter went to look at the house that the wind destroyed.
\end{verbatim}
\end{frame}

@@

\small
\begin{verbatim}
>>> retain = ('lx', 'ps', 'ge')
>>> for entry in lexicon:
...     entry[:] = [f for f in entry if f.tag in retain]
>>> for field in lexicon[50]:
...     print "\\%s %s" % (field.tag, field.text)
\lx kaeviro
\ps V.A
\ge lift off
\ge take off
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> def num_syls(word):
...     template = cv(word)
...     num_cons = template.count('C')
...     return num_cons
>>> def is_vowel(char):
...     return (char in 'aeiou')
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> for entry in lexicon:
...     lx = entry.findtext('lx')
...     if lx:
...         ps = entry.findtext('ps')
...         if num_syls(lx) == 3 and ps[0] == 'V'\
...           and is_vowel(lx[-1]) and is_vowel(lx[-2]):
...             ge = entry.findtext('ge')
...             print "%s (%s) '%s'" % (lx, ps, ge)
kaetupie (V.B) 'tighten'
kakupie (V.B) 'shout'
kapatau (V.B) 'add to'
kapuapie (V.B) 'wound'
kapupie (V.B) 'close tight'
kapuupie (V.B) 'close'
karepie (V.B) 'return'
karivai (V.A) 'have an appetite'
kasipie (V.B) 'care for'
kaukaupie (V.B) 'shine intensely'
kavorou (V.A) 'covet'
kavupie (V.B) 'leave.behind'
...
kuverea (V.A) 'incomplete'
\end{verbatim}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Generating Summary Reports}
\begin{itemize}
\item overall picture of the quality and organisation of the data
\item E.g. entry patterns
\end{itemize}

>>>>

\begin{frame}
\frametitle{Tools and Technologies}
\begin{itemize}
\item General-purpose tools
  \begin{itemize}
  \item word processors
  \item spreadsheets
  \item databases
  \end{itemize}
\item Special purpose tools
  \begin{itemize}
  \item Toolbox
  \end{itemize}
\item broader question: data management in support of best practices
\item Bird, Steven and Gary Simons (2003).  Seven Dimensions of Portability
    for Language Documentation and Description, \textit{Language} 79: 557-582.
\end{itemize}
\end{frame}

@@

\begin{verbatim}
<p class=MsoNormal>
  sleep
  <span style='mso-spacerun:yes'> </span>
  [<span class=SpellE>sli:p</span>]
  <span style='mso-spacerun:yes'> </span>
  <b><span style='font-size:11.0pt'>vi</span></b>
  <span style='mso-spacerun:yes'> </span>
  <i>a condition of body and mind ...</i>
</p>
\end{verbatim}

@@

\begin{itemize}
\item scan the file, convert into tree object
\item preserves order of fields, gives array and XPath-style access
\end{itemize}

\begin{verbatim}
>>> from nltk_lite.corpora import toolbox
>>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
\end{verbatim}
\end{frame}

@@

\scriptsize
\begin{verbatim}
>>> toolbox.to_sfm_string(lexicon[3])
\lx kaa
\ps N.M
\cl isi
\ge cooking banana
\gp banana bilong kukim
\sf FLORA
\dt 12/Feb/2005
\ex Taeavi iria kaa isi kovopaueva kaparapasia.
\xp Taeavi i bin planim gaden banana bilong kukim tasol long paia.
\xe Taeavi planted banana in order to cook it.
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> from nltk_lite.etree.ElementTree import SubElement
>>> for entry in lexicon:
...     for field in entry:
...	    if field.tag == 'lx':
...         cv_field = SubElement(entry, 'cv')
...         cv_field.text = cv(field.text)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Adding New Fields (cont)}
\small
\begin{verbatim}
>>> toolbox.to_sfm_string(lexicon[50])
\lx kaeviro
\cv CVVCVCV
\ps V.A
\ge lift off
\ge take off
\gp go antap
\nt used to describe action of plane
\dt 12/Feb/2005
\ex Pita kaeviroroe kepa kekesia oa vuripierevo kiuvu.
\xp Pita i go antap na lukim haus win i bagarapim.
\xe Peter went to look at the house that the wind destroyed.
\end{verbatim}
\end{frame}

@@

\small
\begin{verbatim}
>>> retain = ('lx', 'ps', 'ge')
>>> for entry in lexicon:
...     entry[:] = [f for f in entry if f.tag in retain]
>>> toolbox.to_sfm_string(lexicon[50])
\lx kaeviro
\ps V.A
\ge lift off
\ge take off
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> def num_cons(word):
...     template = cv(word)
...     return template.count('C')
>>> def is_vowel(char):
...     return (char in 'aeiou')
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> for entry in lexicon:
...     lx = entry.findtext('lx')
...     if lx:
...         ps = entry.findtext('ps')
...         if num_cons(lx) == 3 and ps[0] == 'V'\
...           and is_vowel(lx[-1]) and is_vowel(lx[-2]):
...             ge = entry.findtext('ge')
...             print "%s (%s) '%s'" % (lx, ps, ge)
kaetupie (V.B) 'tighten'
kakupie (V.B) 'shout'
kapatau (V.B) 'add to'
kapuapie (V.B) 'wound'
kapupie (V.B) 'close tight'
kapuupie (V.B) 'close'
karepie (V.B) 'return'
karivai (V.A) 'have an appetite'
kasipie (V.B) 'care for'
kaukaupie (V.B) 'shine intensely'
kavorou (V.A) 'covet'
kavupie (V.B) 'leave.behind'
...
kuverea (V.A) 'incomplete'
\end{verbatim}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Generating Summary Reports}
\begin{itemize}
\item overall picture of the quality and organisation of the data
\item e.g. average number of fields per record
\end{itemize}

\begin{verbatim}
>>> sum(len(entry) for entry in lexicon) / len(lexicon)
10
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Generating Summary Reports (cont)}
\begin{itemize}
\item E.g. entry patterns
\end{itemize}
